
---
title: "Project"
author: "Farah Saleem"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
##Introduction: I chose the datasets survey and uggs because they included college student data which was really interesting and relevant to me as a college student myself. They contain variables such as piercing,tattoos and sleeping patterns along with age, height and smoking status. I chose both of these from packages installed in R. 
```{R}
library("MASS")
library(VGAMdata)
library(dplyr)
library(ggplot2)
library(tidyr)
library(tidyverse)
survey1<-survey%>%mutate(Sex=tolower(Sex))
fs<-survey1%>%pivot_wider(names_from="Sex", values_from="Smoke")
survey2<-fs%>%pivot_longer(c("female","male"), names_to="Sex", values_to="Smoke")
fs<-full_join(survey2,ugss,by=c("Age"="age","Sex"="sex"))
fs<-fs%>%dplyr::select(-`NA`)

```

##Tidying: Although the dataset was already tidy, i used pivot_wider to use names from sex (male and female) to become separate columns with values from smoke (smoking status) in these columns. I then undid this by using pivot_longer. I changed the uppercase of Sex in survey to lowercase, so that the data could be joined. 
##Joining: I used full_join for these data sets because this retains all the original data, in case I wanted to create more plots/correlation charts for future purposes. 
```{R}
library(dplyr)
fs%>%filter(Clap=="Left")
fs%>%slice(1:3)                 
fs%>%arrange(Age)
fs%>%dplyr::select(1,2)
fs%>%mutate(average=Wr.Hnd/NW.Hnd)
fs%>% summarize_all(n_distinct)
fs%>% group_by(Clap)%>%summarize(meanage=mean(Age,na.rm=T))
fs%>% group_by(Sex)%>% summarize(meanage=mean(Age,na.rm=T),sdage=sd(Age,na.rm=T),varAge=var(Age,na.rm=T),meanWr.Hnd=mean(Wr.Hnd,na.rm=T),sdWr=sd(Wr.Hnd,na.rm=T),varwr=var(Wr.Hnd,na.rm=T),meanNW=mean(NW.Hnd,na.rm=T),sdNW=sd(NW.Hnd, na.rm=T),varNW=var(NW.Hnd,na.rm=T),mnNW=min(NW.Hnd,na.rm=T))
library(ggplot2)
```
##Wrangling: Filtering with clap=left, filtered out all the claps apart from those that were left, it also showed that most left clap folded with right hand. Slicing 1:3, sliced out rows 1-3 from the rest of the data.Arranging by age organized the data according to ascending order of age. Select (1,2) allowed selection of the first 2 columns. Mutate (average=Wr.Hnd/NW.Hnd) created another column with the average of wr.hnd values divided by nw.hnd values. summarize_all(n_distinct) summarized all the values that were distinct. Grouping by clap using summarize to create mean age, showed the mean ages in every clap group. Grouping by sex and creating statistics mean,var,sd of age, wr.hnd, nw.hnd showed the result of these statistics in male and female groups separately. 

```{r}



ggplot(fs, aes(x = Pulse, y = Age, fill=Exer))+  geom_bar(alpha = 3/5,stat="summary",fun.y="mean", position="dodge")+  geom_errorbar(stat="summary",position="dodge")+ggtitle("Pulse by Age and Exercise frequence")
```
##This graph shows the relationship of Age and Pulse separated by Exercise. since this is data from college students, the age is usually around 20s, apart from one outlier in his sixties. Those who frequently exercise seem to have a pulse in the middle range around 70, while those who don't exercise at all have a higher unhealthier pulse. Those who exercise some times have a varying range of pulse, from a really low value of around 12 to high around 100. 
```{r}
plot2<-ggplot(na.omit(fs)) + 
  geom_point(mapping = aes(x = Age, y = Smoke),na.rm=TRUE) + 
  facet_wrap(~Sex, nrow = 2)
plot2+ggtitle("Smoking frequency in male and females by Age")
```

##This plot shows that none of the females smoked, so more males smoked than females. It also showed that ages 20 and younger smoke more than those above 20, and compared to regular smokers, there are more males that do not smoke at all. 


```{r}
fs1<-fs%>%dplyr::select(-piercings,-pierced,-tattoos,-glasses)
fs_nums<-fs1%>%select_if(is.numeric)%>%scale
rownames(fs_nums)<-fs$Exer
fs_pca<-princomp(na.omit(fs_nums), cor = TRUE, scores= TRUE)
names(fs_pca)
summary(fs_pca,loadings=T)
eigval<-fs_pca$sdev^2 
varprop=round(eigval/sum(eigval),2)
ggplot()+geom_bar(aes(y=varprop,x=1:17),stat="identity")+xlab("")+geom_path(aes(y=varprop,x=1:17))+  geom_text(aes(x=1:17,y=varprop,label=round(varprop,2)),vjust=1,col="white",size=5)+  scale_y_continuous(breaks=seq(0,.6,.2),labels = scales::percent)+  scale_x_continuous(breaks=1:10)
round(cumsum(eigval)/sum(eigval),2)
eigval
ggplot()+geom_point(aes(fs_pca$scores[,1], fs_pca$scores[,2]))+xlab("PC1")+ylab("PC2")
fs_pca$loadings[1:7,1:2]%>%as.data.frame%>%rownames_to_column%>%
  ggplot()+geom_hline(aes(yintercept=0),lty=2)+
  geom_vline(aes(xintercept=0),lty=2)+ylab("PC2")+xlab("PC1")+
  geom_segment(aes(x=0,y=0,xend=Comp.1,yend=Comp.2),arrow=arrow(),col="red")+
  geom_label(aes(x=Comp.1*1.1,y=Comp.2*1.1,label=rowname))
```

##The first graph shows that pc1 and pc2 count for 0.27 of the total variance.The 2nd graph depends on the angle between the vectors, hence it shows that height and study are negatively correlated, while W.hnd,NW.Hnd and height are almost reduntant. PC1 and PC2 are not directly correlated because the increase of pc1 does not affect the increase pf pc2. 
